canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada's largest metropolitan areas.

### 2025-04-07 paper/Manuscript-Data-Package.tex
fixed spelling errors
package name (canaccessR); data-based to data-driven; OPDs to ODP

\begin{abstract}
In this paper, we describe the \{canaccesR\} package, an open data
product (ODP) created in R that contains public transit travel time
estimates to employment locations and grocery stores across Canada's 12
largest metropolitan areas. We calculate travel time matrices (TTM) from
and to each Dissemination Area (DA) within these regions for the years
2019 and 2023. We add value to the urban analytics community by
processing and integrating raw data, and disseminating user-ready data
in the domain of transportation accessibility in Canada. To do so, we
use the \{r5r\} R package, General Transit Feed Specification (GTFS),
OpenStreetMap (OSM), DMTI's Enhanced Points of Interest, and Statistics
Canada Census data. This data package can be used by researchers,
practitioners, and transit agencies to estimate accessibility levels to
these two essential destinations within these urban areas. Moreover,
these estimations can be used as inputs in equity and inequalities
assessments, through the comparisons of within and across accessibility
levels found throughout Canada's largest metropolitan areas.
Consequently, we expect to contribute to informed and data-based
decision making in transportation by disseminating these data. We hope
that these datasets can substantiate future improvements in
policy-making that may lead to greater justice in the country's urban
transportation systems. The package is still in its initial phase and
may undergo expansions in the future by adding TTM's for other
destinations (e.g., schools, healthcare facilities). Finally, as an
ODPs, the \{canaccess\} package allows for open exploration, use, and
contribution by users through its GitHub repository.
\end{abstract}

@@

The objective of this paper is to describe the \{canaccesR\} open data
package. Its main contents are a set of public transit travel time
matrices (TTM) estimates to employment and groceries stores from the 12
largest Canadian metropolitan areas in 2019 and 2023, representing
approximately 55\% of the Canadian population. The results are provided
at the Dissemination Areas (DA) \footnote{Dissemination Areas are the
  smallest publicly available spatial unit provided by Statistics Canada
  \citep{governmentofcanadaDictionaryCensusPopulation2021a}.} level,
yielding origin-destination pairs containing associated travel time by
public transit, population, total employment, mode share, and other
relevant census variables and spatial shape boundaries for each
metropolitan area. This data package was created by leveraging expertise
in data science, R programming, and transportation analysis. It includes
.Rmd notebooks written for computing travel time matrices for large sets
of origin-destination pairs using the \{r5r\} R package
\citep{pereiraR5rRapidRealistic2021}. Overall, \{canaccesR\} offers 52
complementary objects ready for temporal and spatial analysis. The
package is an analysis-read product based on a fusion of data sources,
including public transit schedules (GTFS files), road and transit
networks, census data, and filtered business location data.

TTMs are a core piece of information required for estimating spatial
interaction. Accessibility -the \emph{potential} spatial interaction
offered by the transportation system to reach destinations
\citep{paezMeasuringAccessibilityPositive2012}- is an example of a
measure that requires TTMs. Recent efforts following an open-source and
transparent philosophy have been made to disseminate useful data and
information on transportation in the Canadian context
\citep{soukhovTTS2016RDataSet2023}. However, despite these initiatives,
pre-processed and available data that allow for the ease estimation of
accessibility indicators are still scarce. Within this context, we
expect to help filling this gap by the processing raw into user-ready
data and making them publicly available to advance knowledge on the
field. Our main contribution is to provide analysis-ready data for
Canada's largest cities on the topic of transportation accessibility,
thus making urban analytics in the country more accessible and
contributing to future research and data-based decision making.

The package's main audiences are Canadian researchers in urban planning
and transportation and transportation system agencies. We anticipate
three primary uses for the open data product (ODP) described in this
paper. First, the datasets allow for static assessment of the level of
public transit accessibility across the country's largest cities before
and after the COVID-19 pandemic. In other words, \{canaccesR\} makes it
easier for those interested in comparing cities regarding their level of
public transit accessibility to essential destinations (such as
employment centers and groceries stores) to do so. Second, the temporal
and spatial characters of the datasets made available here allow
researchers to evaluate accessibility changes through time and across
space within the largest Canadian urban areas. Third, as is now common
practice in transportation accessibility research, used as inputs, these
estimates can substantiate broader investigations on transportation
justice and equity
\citep{higginsChangesAccessibilityEmergency2021, humbertoHowTranslateJustice2023, pereiraGeographicAccessCOVID192021}.
For example, the TTM estimates allow for evaluating the evolution of
public transit's accessibility by income or spatial distribution across
all Dissemination Areas (DA's) of each of the 12 cities in the sample
\citep{pargaDemocraticAccessOur2024}. In other words, the package's
contents can be used from straightforward assessments of accessibility
in Canadian urban areas to more theoretically and morally complex
evaluations of justice in the country's urban transportation system.

Besides this introduction, we organize this paper as follows. The next
section contains a description of the data sources we used to construct
the data package. Then, we recount the data processing necessary to
create the package. Next, we go through the main contents of the data
package, i.e., the travel time matrices estimated through our analysis.
We present some basic descriptive statistics of these datasets, and
elucidate how one can use them in accessibility analysis. Finally, we
conclude by explaining how we expect \{canaccesR\} to contribute to the
urban analytics and science community.

@@

The locations included in the data package comprise the 12 largest
(population-wise) Census metropolitan areas (CMA's) based on the 2016
Canadian Census \citep{governmentofcanada2016CensusPopulation2016}
\footnote{We included Oshawa as part of the Greater Toronto Area (GTA)
  because of its proximity. Similarly, we included Abbotsford-Mission as
  part of the Vancouver metropolitan area due to its proximity to a
  transit station on the region's West Coast Express commuter rail line.}.These
locations are the surrunding Toronto, Montreal, Vancouver,
Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
Kitchener-Cambridge-Waterloo, London, and Halifax areas. We used four
main data sources to construct the \{canaccesR\} data package: General
Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
Points of Interest, and Statistics Canada Census data.

@@

Using the \{r5r\} R package, we estimated public transit travel times
for two destination types, grocery stores and jobs. For each amenity
type, we chose a likely travel time and day of the week. We set a 15
minutes time window and the maximum trip duration to 120 minutes. The
estimated times are the median of the 15 minute time window. For
groceries stores, we set the departure date to a weekend afternoon and
the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and
April 22, 2023. For employment, we ran the analysis on a typical weekday
morning rush-hour commute, more specifically 8:00 to 8:15 AM departure
on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 \footnote{The one
  exception is Quebec City, where the routing for 2019 occurs on a
  Saturday and Tuesday in June (instead of April) due to the GTFS data
  unavailability.}. In both cases, we assumed that walking was the mode
of travel from origin to transit stop and from transit stop to
destination. We aggregated all the resulting travel time matrices at the
DA level, which comprise the fundamental unit of analysis in data
package.

@@

\begin{table}[!t]
\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{1\linewidth}{@{\extracolsep{\fill}}clrrrrrr}
\toprule
Study Region Name & Destination & Observations & Mean & Sd & P25 & P50 & P75 \\ 
\midrule\addlinespace[2.5pt]
All regions & Employment & 97,784,850 & 78 & 25 & 59 & 80 & 99 \\ 
Toronto & Employment & 41,038,062 & 82 & 25 & 64 & 85 & 103 \\ 
Montréal & Employment & 30,091,411 & 76 & 25 & 57 & 77 & 97 \\ 
Vancouver & Employment & 12,254,478 & 76 & 25 & 58 & 77 & 96 \\ 
Calgary & Employment & 3,022,244 & 75 & 23 & 59 & 75 & 91 \\ 
Ottawa & Employment & 2,936,912 & 74 & 24 & 56 & 74 & 93 \\ 
Edmonton & Employment & 2,057,486 & 70 & 23 & 53 & 70 & 87 \\ 
Québec City & Employment & 1,473,702 & 72 & 25 & 54 & 72 & 93 \\ 
Winnipeg & Employment & 1,574,205 & 62 & 22 & 46 & 61 & 76 \\ 
Hamilton & Employment & 2,089,172 & 82 & 28 & 61 & 87 & 106 \\ 
Waterloo & Employment & 552,826 & 71 & 27 & 49 & 69 & 93 \\ 
London & Employment & 415,174 & 61 & 21 & 46 & 60 & 74 \\ 
Halifax & Employment & 279,178 & 66 & 25 & 47 & 65 & 84 \\ 
All regions & Groceries Stores & 18,519,897 & 76 & 26 & 56 & 76 & 97 \\ 
Toronto & Groceries Stores & 8,512,874 & 80 & 25 & 61 & 82 & 101 \\ 
Montréal & Groceries Stores & 2,993,965 & 73 & 26 & 53 & 74 & 94 \\ 
Vancouver & Groceries Stores & 4,540,106 & 73 & 25 & 53 & 74 & 93 \\ 
Calgary & Groceries Stores & 502,757 & 70 & 22 & 54 & 70 & 86 \\ 
Ottawa & Groceries Stores & 619,791 & 72 & 24 & 55 & 72 & 91 \\ 
Edmonton & Groceries Stores & 324,030 & 68 & 23 & 51 & 67 & 84 \\ 
Québec City & Groceries Stores & 234,600 & 70 & 26 & 51 & 70 & 89 \\ 
Winnipeg & Groceries Stores & 362,566 & 58 & 21 & 43 & 57 & 72 \\ 
Hamilton & Groceries Stores & 281,118 & 84 & 29 & 60 & 91 & 110 \\ 
Waterloo & Groceries Stores & 44,726 & 66 & 28 & 44 & 62 & 90 \\ 
London & Groceries Stores & 52,617 & 55 & 19 & 42 & 55 & 67 \\ 
Halifax & Groceries Stores & 50,747 & 63 & 26 & 42 & 63 & 83 \\ 
\bottomrule
\end{tabular*}
\end{table}

\section{How to use \{canaccessR\}}\label{how-to-use-canaccessr}

This section exemplifies how the package can be used through visual
representation. In Figure \citet{fig-travel_time_emp_grc_plot}, we
present the spatial representation of the travel time matrices for the
metropolitan region of Montréal. In it, we see the median travel time
from each DA to employment (left) and to groceries stores (right). The
plot shows that moving away from the city core increases the necessary
travel time by public transit to reach employment locations and
groceries stores.

\begin{figure}[H]
\includegraphics[width=1\linewidth]{../figures/patch_tt_emp_grc} \caption{Estimated median travel time (minutes) per Dissemination Area to jobs (left) and groceries stores (right). Public transit travel times are calculated using {r5r} (Pereira et al., 2021). DA and planning boundaries of the Montréal metropolitan area (Statistics Canada, 2016).}\label{fig:fig-travel_time_emp_grc_plot}
\end{figure}

@@

In this paper, we describe the \{canaccesR\} data package, created using
the \{r5r\} package and transit schedule, street network, employment,
and population data. The package's main contents refers to the
ready-to-use travel time matrices for public transit to reach employment
and groceries stores in Canada's 12 largest urban areas. We expect the
contents of the package to be used in transportation accessibility
evaluations within and across those regions. Moreover, these datasets
can be used in further equity assessments that evaluate the distribution
of accessibility across space and between social groups. Furthermore, in
the spirit of open data products
\citep{arribas-belOpenDataProductsA2021}, the package can be expanded
through collaboration with other researchers by, for example, including
travel time matrices to other essential destinations within the DMTI's
dataset (\emph{e.g.}, schools, healthcare, etc.). In other words, we
hope that by making these datasets publicly available, future analysis
can contribute to making Canada's transportation system more just and
fair, considering accessibility's as the main social good of
transportation \citep{martensTransportJusticeDesigning2016}, and the
inherent connection between public transit and the ``right to the city''
\citep{cogginRightTransportMoving2015}.

>>>>

\begin{abstract}
In this paper, we describe the \{canaccessR\} package, an open data
product (ODP) created in R that contains public transit travel time
estimates to employment locations and grocery stores across Canada's 12
largest metropolitan areas. We calculate travel time matrices (TTM) from
and to each Dissemination Area (DA) within these regions for the years
2019 and 2023. We add value to the urban analytics community by
processing and integrating raw data, and disseminating user-ready data
in the domain of transportation accessibility in Canada. To do so, we
use the \{r5r\} R package, General Transit Feed Specification (GTFS),
OpenStreetMap (OSM), DMTI's Enhanced Points of Interest, and Statistics
Canada Census data. This data package can be used by researchers,
practitioners, and transit agencies to estimate accessibility levels to
these two essential destinations within these urban areas. Moreover,
these estimations can be used as inputs in equity and inequalities
assessments, through the comparisons of within and across accessibility
levels found throughout Canada's largest metropolitan areas.
Consequently, we expect to contribute to informed and data-driven
decision making in transportation by disseminating these data. We hope
that these datasets can substantiate future improvements in
policy-making that may lead to greater justice in the country's urban
transportation systems. The package is still in its initial phase and
may undergo expansions in the future by adding TTM's for other
destinations (e.g., schools, healthcare facilities). Finally, as an ODP,
the \{canaccess\} package allows for open exploration, use, and
contribution by users through its GitHub repository.
\end{abstract}

@@

The objective of this paper is to describe the \{canaccessR\} open data
package. Its main contents are a set of public transit travel time
matrices (TTM) estimates to employment and groceries stores from the 12
largest Canadian metropolitan areas in 2019 and 2023, representing
approximately 55\% of the Canadian population. The results are provided
at the Dissemination Areas (DA) \footnote{Dissemination Areas are the
  smallest publicly available spatial unit provided by Statistics Canada
  \citep{governmentofcanadaDictionaryCensusPopulation2021a}.} level,
yielding origin-destination pairs containing associated travel time by
public transit, population, total employment, mode share, and other
relevant census variables and spatial shape boundaries for each
metropolitan area. This data package was created by leveraging expertise
in data science, R programming, and transportation analysis. It includes
.Rmd notebooks written for computing travel time matrices for large sets
of origin-destination pairs using the \{r5r\} R package
\citep{pereiraR5rRapidRealistic2021}. Overall, \{canaccessR\} offers 52
complementary objects ready for temporal and spatial analysis. The
package is an analysis-read product based on a fusion of data sources,
including public transit schedules (GTFS files), road and transit
networks, census data, and filtered business location data.

TTMs are a core piece of information required for estimating spatial
interaction. Accessibility -the \emph{potential} spatial interaction
offered by the transportation system to reach destinations
\citep{paezMeasuringAccessibilityPositive2012}- is an example of a
measure that requires TTMs. Recent efforts following an open-source and
transparent philosophy have been made to disseminate useful data and
information on transportation in the Canadian context
\citep{soukhovTTS2016RDataSet2023}. However, despite these initiatives,
pre-processed and available data that allow for the ease estimation of
accessibility indicators are still scarce. Within this context, we
expect to help filling this gap by the processing raw into user-ready
data and making them publicly available to advance knowledge on the
field. Our main contribution is to provide analysis-ready data for
Canada's largest cities on the topic of transportation accessibility,
thus making urban analytics in the country more accessible and
contributing to future research and data-driven decision making.

The package's main audiences are Canadian researchers in urban planning
and transportation and transportation system agencies. We anticipate
three primary uses for the open data product (ODP) described in this
paper. First, the datasets allow for static assessment of the level of
public transit accessibility across the country's largest cities before
and after the COVID-19 pandemic. In other words, \{canaccessR\} makes it
easier for those interested in comparing cities regarding their level of
public transit accessibility to essential destinations (such as
employment centers and groceries stores) to do so. Second, the temporal
and spatial characters of the datasets made available here allow
researchers to evaluate accessibility changes through time and across
space within the largest Canadian urban areas. Third, as is now common
practice in transportation accessibility research, used as inputs, these
estimates can substantiate broader investigations on transportation
justice and equity
\citep{higginsChangesAccessibilityEmergency2021, humbertoHowTranslateJustice2023, pereiraGeographicAccessCOVID192021}.
For example, the TTM estimates allow for evaluating the evolution of
public transit's accessibility by income or spatial distribution across
all Dissemination Areas (DA's) of each of the 12 cities in the sample
\citep{pargaDemocraticAccessOur2024}. In other words, the package's
contents can be used from straightforward assessments of accessibility
in Canadian urban areas to more theoretically and morally complex
evaluations of justice in the country's urban transportation system.

Besides this introduction, we organize this paper as follows. The next
section contains a description of the data sources we used to construct
the data package. Then, we recount the data processing necessary to
create the package. Next, we go through the main contents of the data
package, i.e., the travel time matrices estimated through our analysis.
We present some basic descriptive statistics of these datasets, and
elucidate how one can use them in accessibility analysis. Finally, we
conclude by explaining how we expect \{canaccessR\} to contribute to the
urban analytics and science community.

@@

The locations included in the data package comprise the 12 largest
(population-wise) Census metropolitan areas (CMA's) based on the 2016
Canadian Census \citep{governmentofcanada2016CensusPopulation2016}
\footnote{We included Oshawa as part of the Greater Toronto Area (GTA)
  because of its proximity. Similarly, we included Abbotsford-Mission as
  part of the Vancouver metropolitan area due to its proximity to a
  transit station on the region's West Coast Express commuter rail line.}.These
locations are the surrunding Toronto, Montreal, Vancouver,
Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
Kitchener-Cambridge-Waterloo, London, and Halifax areas. We used four
main data sources to construct the \{canaccessR\} data package: General
Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
Points of Interest, and Statistics Canada Census data.

@@

Using the \{r5r\} R package, we estimated public transit travel times
for two destination types, grocery stores and jobs. For each amenity
type, we chose a likely travel time and day of the week. We set a 15
minutes time window and the maximum trip duration to 120 minutes. The
estimated times are the median of the 15 minute time window. For
groceries stores, we set the departure date to a weekend afternoon and
the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and
April 22, 2023. For employment, we ran the analysis on a typical weekday
morning rush-hour commute, more specifically 8:00 to 8:15 AM departure
on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 \footnote{The one
  exception is Quebec City, where the routing for 2019 occurs on a
  Saturday and Tuesday in June (instead of April) due to the GTFS data
  unavailability.}. In both cases, we assumed that walking was the mode
of travel from origin to transit stop and from transit stop to
destination. We aggregated all the resulting travel time matrices at the
DA level, which comprise the fundamental spatial unit of analysis in
data package.

@@

\begingroup
\setlength\LTleft{0\linewidth}
\setlength\LTright{0\linewidth}\fontsize{7.5pt}{9.0pt}\selectfont
\begin{longtable}{@{\extracolsep{\fill}}clrrrrrr}
\toprule
Study Region Name & Destination & Observations & Mean & Sd & P25 & P50 & P75 \\ 
\midrule\addlinespace[2.5pt]
All regions & Employment & 97,784,850 & 78 & 25 & 59 & 80 & 99 \\ 
Toronto & Employment & 41,038,062 & 82 & 25 & 64 & 85 & 103 \\ 
Montréal & Employment & 30,091,411 & 76 & 25 & 57 & 77 & 97 \\ 
Vancouver & Employment & 12,254,478 & 76 & 25 & 58 & 77 & 96 \\ 
Calgary & Employment & 3,022,244 & 75 & 23 & 59 & 75 & 91 \\ 
Ottawa & Employment & 2,936,912 & 74 & 24 & 56 & 74 & 93 \\ 
Edmonton & Employment & 2,057,486 & 70 & 23 & 53 & 70 & 87 \\ 
Québec City & Employment & 1,473,702 & 72 & 25 & 54 & 72 & 93 \\ 
Winnipeg & Employment & 1,574,205 & 62 & 22 & 46 & 61 & 76 \\ 
Hamilton & Employment & 2,089,172 & 82 & 28 & 61 & 87 & 106 \\ 
Waterloo & Employment & 552,826 & 71 & 27 & 49 & 69 & 93 \\ 
London & Employment & 415,174 & 61 & 21 & 46 & 60 & 74 \\ 
Halifax & Employment & 279,178 & 66 & 25 & 47 & 65 & 84 \\ 
All regions & Groceries Stores & 18,519,897 & 76 & 26 & 56 & 76 & 97 \\ 
Toronto & Groceries Stores & 8,512,874 & 80 & 25 & 61 & 82 & 101 \\ 
Montréal & Groceries Stores & 2,993,965 & 73 & 26 & 53 & 74 & 94 \\ 
Vancouver & Groceries Stores & 4,540,106 & 73 & 25 & 53 & 74 & 93 \\ 
Calgary & Groceries Stores & 502,757 & 70 & 22 & 54 & 70 & 86 \\ 
Ottawa & Groceries Stores & 619,791 & 72 & 24 & 55 & 72 & 91 \\ 
Edmonton & Groceries Stores & 324,030 & 68 & 23 & 51 & 67 & 84 \\ 
Québec City & Groceries Stores & 234,600 & 70 & 26 & 51 & 70 & 89 \\ 
Winnipeg & Groceries Stores & 362,566 & 58 & 21 & 43 & 57 & 72 \\ 
Hamilton & Groceries Stores & 281,118 & 84 & 29 & 60 & 91 & 110 \\ 
Waterloo & Groceries Stores & 44,726 & 66 & 28 & 44 & 62 & 90 \\ 
London & Groceries Stores & 52,617 & 55 & 19 & 42 & 55 & 67 \\ 
Halifax & Groceries Stores & 50,747 & 63 & 26 & 42 & 63 & 83 \\ 
\bottomrule
\end{longtable}
\endgroup

\section{\{canaccessR\} usage}\label{canaccessr-usage}

This section presents an usage example of the package through visual
representation of its data. In Figure
\citet{fig-travel_time_emp_grc_plot}, we show the spatial distribution
of the travel time matrices for the metropolitan region of Montréal. In
it, we see the median travel time from each DA to employment (left) and
to groceries stores (right). The plot shows that moving away from the
city core increases the necessary travel time by public transit to reach
employment locations and groceries stores.

\begin{figure}[H]
\includegraphics[width=1\linewidth]{../figures/patch_tt_emp_grc} \caption{Estimated median travel time (minutes) per Dissemination Area to jobs (left) and groceries stores (right). Public transit travel times are calculated using {r5r} (Pereira et al., 2021), and DA and planning boundaries of the Montréal metropolitan area use Statistics Canada data (Statistics Canada, 2016).}\label{fig:fig-travel_time_emp_grc_plot}
\end{figure}

@@

In this paper, we describe the \{canaccessR\} data package, created
using the \{r5r\} package and transit schedule, street network,
employment, and population data. The package's main contents refers to
the ready-to-use travel time matrices for public transit to reach
employment and groceries stores in Canada's 12 largest urban areas. We
expect the contents of the package to be used in transportation
accessibility evaluations within and across those regions. Moreover,
these datasets can be used in further equity assessments that evaluate
the distribution of accessibility across space and between social
groups. Furthermore, in the spirit of open data products
\citep{arribas-belOpenDataProductsA2021}, the package can be expanded
through collaboration with other researchers by, for example, including
travel time matrices to other essential destinations within the DMTI's
dataset (\emph{e.g.}, schools, healthcare, etc.). In other words, we
hope that by making these datasets publicly available, future analysis
can contribute to making Canada's transportation system more just and
fair, considering accessibility's as the main social good of
transportation \citep{martensTransportJusticeDesigning2016}, and the
inherent connection between public transit and the ``right to the city''
\citep{cogginRightTransportMoving2015}.
